% =====================================================================
% REMORA: A Policy-Gated Multi-Oracle Assurance Architecture for Agentic AI
% LaTeX source — suitable for Overleaf, NeurIPS/ICML/ACL workshop templates
% =====================================================================
\documentclass[10pt,twocolumn]{article}

% ---- Packages -------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{hyperref}
% xurl = url with break points after every character class, so long
% URLs and \path tokens wrap inside the narrow two-column measure
% instead of protruding into the margin.
\usepackage{xurl}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{listings}
% \usepackage{algorithm}    % not installed in build env; no algorithm blocks used
% \usepackage{algpseudocode}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{tcolorbox}

% ---- Code listings ---------------------------------------------------
% Global listings setup. verbatim must not be used in this paper: it
% never breaks lines, so any long line silently overflows the column in
% twocolumn mode. lstlisting with breaklines wraps instead.
\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  breakatwhitespace=false,
  postbreak=\mbox{\textcolor{gray}{$\hookrightarrow$}\space},
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  framerule=0.3pt,
  rulecolor=\color{black!25},
  xleftmargin=4pt,
  xrightmargin=4pt,
  aboveskip=6pt,
  belowskip=6pt,
}

% Fit a tabular to the column width. Used for tables whose natural width
% exceeds \columnwidth in twocolumn mode; scaling preserves the layout
% without overflowing into the neighbouring column.
\newcommand{\fitcolumn}[1]{\resizebox{\columnwidth}{!}{#1}}

% Paragraph-level overflow tolerance for narrow two-column measure:
% allow modest extra inter-word stretch before a line is allowed to
% protrude. Removes residual overfull lines (long URLs, technical
% tokens) without visibly loosening the typography.
\setlength{\emergencystretch}{2em}
\tolerance=2000

% ---- Theorem environments -------------------------------------------
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}

% ---- Metadata -------------------------------------------------------
\title{REMORA: A Policy-Gated Multi-Oracle\\Assurance Architecture for Agentic AI}

% Paper version — keep in lockstep with the repository release tag.
\newcommand{\paperversion}{v0.9.0}

\author{
  Stian Skogbrott\thanks{Developer and maintainer of REMORA. Luftfiber AS.
  Contact: \texttt{support@luftfiber.no}.}\\
  Luftfiber AS\\
  \texttt{https://github.com/darklordVirtual/REMORA}
}

\date{\paperversion\ --- content dated June 2026 \ \small compiled \today\ by CI from the repository LaTeX source (unchanged content)}

% ---- Notation -------------------------------------------------------
\newcommand{\Oracles}{\mathcal{O}}
\newcommand{\OraclesV}{\mathcal{O}'}
\renewcommand{\phi}{\varphi}
\newcommand{\phat}{\hat{p}}
\newcommand{\trust}{\tau}
\newcommand{\lyap}{V}
\newcommand{\gate}{g}
\newcommand{\Gate}{\Gamma}

% =====================================================================
\begin{document}

\maketitle

% =====================================================================
\begin{abstract}
Autonomous AI agents that invoke tools, query databases, or actuate
real-world systems require an assurance layer that decides---before
execution---whether an action is safe enough to proceed autonomously,
requires verification, warrants abstention, or must be escalated to
human review. We present \textbf{REMORA} (Reasoning Ensemble
Multi-Oracle Routing Architecture), a research-grade control layer for
agentic AI governance. REMORA combines parallel multi-oracle consensus,
canonicalized verdict extraction, correlation-aware consensus
weighting, thermodynamic-inspired uncertainty observables,
Lyapunov-inspired heuristic stability tracking, and a policy engine with
hard-block precedence rules to produce one of four structured gate
outcomes: \textsc{accept}, \textsc{verify}, \textsc{abstain}, or
\textsc{escalate}.

On a 544-item benchmark (TruthfulQA, BoolQ, adversarial suite), REMORA
achieves 88.8\% selective accuracy at 18\% coverage (majority-vote
full-coverage baseline: 41.18\%; $+47.6$\,pp lift; in-sample optimum).
Exploiting a preliminary critical-phase trust inversion observation ($N=32$ real-oracle
critical items; low-$\tau$
critical items achieve 71.4\% accuracy vs.\ 27.3\% for high-$\tau$),
a \texttt{PhaseAwareGuardrail} with inverted-score selection extends coverage
to 22.1\% at 85.0\% accuracy ($+43.8$\,pp lift; $N=120$; Wilson~CI [77.5\%,
90.3\%])---a $+3.9$\,pp coverage gain with only $1.9$\,pp accuracy cost.
A stratified held-out evaluation ($\tau^*$ locked from training split) confirms
88.0\% accuracy at 23.2\% holdout coverage ($N_\text{accepted}=25$;
$p=1.45\times10^{-5}$; Wilson~CI [70.0\%, 95.8\%]---the wide interval reflects the
small accepted set and should be read as a directional confirmation, not a
tight estimate). On an adversarial agentic tool-call benchmark ($N=700$),
REMORA's full policy gate reduces unsafe execution from 10--20\% (baselines)
to 0\% on benchmark tasks (unsafe-rate Wilson CI [0.00\%, 0.55\%]) while improving
mean utility from $0.0$ to $0.62$. A critical-phase evidence router resolves
38.5\% of high-uncertainty cases with 100\% precision on a 3,000-item NLI
benchmark ($N_{\text{evidence-accept}}=304$, Wilson CI [98.75\%, 100.00\%]); the
remainder are escalated to human review.

Key limitations: (1) the benchmark contains 13.8\% author-curated
items subject to selection bias; (2) trust scoring alone cannot
discriminate correctness in the critical phase (anticorrelation
confirmed empirically); (3) semantic evidence retrieval is pluggable
but not live in the current implementation; (4) the audit hash-chain
is tamper-evident but not tamper-proof without external append-only
storage; (5) the entropy computation in all reported experiments uses a
token-fingerprint heuristic backend, not full Semantic Entropy over NLI
clusters---the NLI backend exists as a drop-in but was not used here.
An important architectural finding: the unsafe-execution reduction from
10\% to 0\% is attributable entirely to deterministic policy hard blocks;
the multi-oracle thermodynamic machinery contributes to decision routing and
calibration quality, not to the safety guarantee itself.
REMORA is a governed autonomy layer (not a replacement for domain
authority) that converts model disagreement, uncertainty, missing
evidence, and policy triggers into explicit, auditable control
decisions.
\end{abstract}

\textbf{Keywords:} agentic AI safety, multi-oracle consensus, selective
prediction, uncertainty routing, policy-as-code, audit governance,
human-in-the-loop.

% =====================================================================
\section{Introduction}
\label{sec:intro}

The deployment of LLM-powered agents capable of invoking tools
introduces a class of decision problem distinct from conversational AI:
an agent proposes a \emph{concrete action} in the world, and the
system must decide whether to execute it. Unlike a hallucinated
sentence, an erroneously executed database write or wrongly triggered
process shutdown has immediate, potentially irreversible consequences.

Existing safeguards (RLHF alignment, system prompts, content
classifiers) are semantic filters applied at training or prompt time.
They are not designed to evaluate a specific proposed action against a
specific operational context, evidence base, and regulatory policy at
inference time. Majority vote among LLMs improves accuracy but cannot
block a confident, wrong consensus from triggering unsafe execution.

We argue that what is needed is not a \emph{better oracle} but an
\emph{assurance layer}: a system that evaluates whether the conditions
for autonomous action are met before anything is executed. When those
conditions are not met, the system must route toward verification,
abstention, or human review, recording the reasoning in an auditable
envelope.

REMORA demonstrates such a layer. Its core thesis:
\textbf{governed autonomy requires explicit routing of
uncertainty---not suppression of it.}

The contribution of this paper is a \textbf{system architecture and
empirical evaluation for governed agentic AI decision control}---not a
new foundation model, training procedure, or benchmark. It is a
reference implementation, a set of measurable observables, a
documented set of negative results, and an honest account of what
remains unsolved.

% =====================================================================
\section{Background and Related Work}
\label{sec:related}

\subsection{LLM Ensembles and Self-Consistency}
Wang et al.\ (2023) \cite{wang2023self} introduced self-consistency
sampling over chain-of-thought paths. Lakshminarayanan et al.\ (2017)
\cite{lakshminarayanan2017simple} demonstrated that deep ensembles from
distinct random initialisations improve both accuracy and calibration.
REMORA's oracle fan-out is structurally related but uses
\emph{distinct model families} rather than multiple samples from one model,
and weights oracles by \emph{inverse pairwise correlation} rather than
treating them as exchangeable. Kuncheva \& Whitaker (2003)
\cite{kuncheva2003measures} formalised diversity measures for classifier
ensembles; REMORA's inverse-correlation weighting can be understood as an
instance of this family applied to instruction-following LLMs.

\subsection{Multi-Agent Debate}
Du et al.\ (2023) \cite{du2023improving} showed that inter-model
debate can reduce hallucination. REMORA's dissensus metric $D$ is a
one-shot, non-iterative analogue: disagreement is measured once, not
resolved through iterative exchange.

\subsection{LLM-as-Judge and Verifier Models}
Zheng et al.\ (2023) \cite{zheng2023judging} established the
LLM-as-judge paradigm. Cobbe et al.\ (2021) \cite{cobbe2021training}
trained process reward models as verifiers. REMORA differs by using a \emph{structured policy engine} rather than a
learned verifier, enabling interpretable hard blocks that cannot be
overridden by a confident oracle.

\subsection{Selective Prediction and Abstention}
Chow (1970) \cite{chow1970optimum} established the optimal error-reject
tradeoff for binary classifiers. El-Yaniv \& Wiener (2010)
\cite{elyaniv2010selective} extended this to a distribution-free
selective classification framework; their coverage--accuracy bound is
the foundational result that Geifman \& El-Yaniv (2017)
\cite{geifman2017selective} apply to deep networks. Kadavath
et al.\ (2022) \cite{kadavath2022language} show that LLM
self-uncertainty estimates are unreliable under distribution shift.
REMORA's abstention is grounded in structural oracle disagreement
rather than self-reported confidence, which aligns with the observation
in \cite{elyaniv2010selective} that selection rules based on model-reported
softmax scores are insufficient under covariate shift.

\subsection{Conformal Prediction}
Angelopoulos \& Bates (2021) \cite{angelopoulos2021gentle} provide a
distribution-free framework for coverage guarantees. REMORA implements
Mondrian phase-stratified conformal calibration for the ordered phase,
and Conformal Risk Control \cite{angelopoulos2022crc} with
importance weights for the critical phase
(\S\ref{sec:crc}). The exchangeability violation in the critical phase
is documented as a negative result (\S\ref{sec:negative}).

\subsection{Prover-Verifier Games}
Kirsch et al.\ (2024) \cite{kirsch2024prover} showed that prover-verifier
games improve LLM output legibility by requiring the prover to justify
claims to an independent verifier.  REMORA adapts this to the offline
multi-oracle case, using Semantic Entropy clustering to identify the prover
(dominant-cluster oracle) and verifier (independent challenger) without
additional API calls (\S\ref{sec:pvd}).

\subsection{AI Governance and Policy-as-Code}
The EU AI Act (2024) \cite{euaiact2024} mandates human oversight and
audit trails for high-risk AI systems. Open Policy Agent (OPA)
\cite{opa} provides a production-grade Rego policy engine. REMORA
integrates an OPA adapter---failing closed to a Python fallback---as a
concrete instantiation of policy-as-code for AI decisions.

\subsection{Assurance Cases}
Kelly (1998) \cite{kelly1998goal} and Bloomfield \& Bishop (2010)
\cite{bloomfield2010safety} structure safety arguments as
claim-evidence-reasoning trees. REMORA's DecisionEnvelope is designed
to populate one node of such a case with a specific gate decision's
evidence chain.

\subsection{Human-in-the-Loop / Human-on-the-Loop}
Endsley (1995) \cite{endsley1995toward} defines supervisory control
levels. REMORA's four-outcome schema directly implements this
spectrum: \textsc{accept} (human-on-the-loop) through
\textsc{escalate} (human-in-the-loop), with \textsc{verify} and
\textsc{abstain} as intermediate states.

% =====================================================================
\section{Problem Statement}
\label{sec:problem}

Let $\mathcal{A}$ be a space of agent-proposed actions, $q$ a query,
and $c = (\text{domain}, \text{risk}, \text{env})$ an operational
context. We seek a control function:
%
\begin{equation}
\begin{split}
\Gate: (q, a, c) \;\longrightarrow\; g, \qquad\qquad\\
g \in \{\textsc{Accept}, \textsc{Verify}, \textsc{Abstain}, \textsc{Escalate}\}
\end{split}
\end{equation}
%
with five required properties: (1)~safety-first (hard policy violations
map to \textsc{Escalate} regardless of consensus); (2)~calibration-aware
(abstain when uncertainty exceeds threshold); (3)~evidence-sensitive
(evidence can resolve or contradict ambiguous consensus);
(4)~auditable (every call produces a signed evidence record);
(5)~conservative by default (fail closed toward \textsc{Verify} or \textsc{Escalate} depending on risk tier).

This is not a question of oracle \emph{accuracy}---REMORA does not
know whether action $a$ is correct. It is a question of
\emph{assurance conditions}: whether the conditions that justify
autonomous execution of $a$ are verifiably met.

% =====================================================================
\section{REMORA Architecture}
\label{sec:architecture}

REMORA processes each gate request through six decision stages,
followed by DecisionEnvelope emission (Stage~7).
Figure~\ref{fig:arch} (Mermaid diagram in supplementary) shows the
full pipeline. Stage~7 is the output record, not a decision stage.

\paragraph{Stage 1: Intake \& Risk Classification.}
The system classifies intent domain, risk tier
($\in\{$low, medium, high, critical$\}$), action type, and target
environment. An adversarial admission firewall checks for injection
patterns (``ignore previous instructions'', ``drop table'', etc.) and
fires an immediate \textsc{Escalate} if detected.

\paragraph{Stage 2: Oracle Fan-Out.}
$n \geq 3$ oracle models from distinct families are queried in parallel
via \texttt{ThreadPoolExecutor}. Responses with non-null
\texttt{error} fields are excluded before any aggregation.
$\OraclesV \subseteq \Oracles$ denotes the valid oracle set.

\paragraph{Stage 3: Canonicalization \& Weighted Consensus.}
Each response is canonicalized by $\varphi: \mathrm{raw} \rightarrow v$
where $v$ is the tuple (\texttt{polarity}, \texttt{claim\_hash},
\texttt{magnitude}, \texttt{tags}).
The \texttt{claim\_hash} field uses a pluggable backend: the default
\texttt{TokenFingerprintBackend} applies a 16-char SHA-256 truncation over
NFKC-normalized sorted tokens for backward-compatible, dependency-free
operation; the \texttt{NLISemanticBackend} replaces this with bidirectional
cross-encoder entailment clustering, implementing Semantic Entropy
\cite{kuhn2023semantic,kuhn2026evse}. Oracle weights are derived from
inverse pairwise correlation (see \S\ref{sec:method}).

\paragraph{Stage 4: Thermodynamic Observables.}
Entropy $H$ is computed over the diversity-weighted verdict distribution
$\hat{p}$. The \texttt{NLISemanticBackend} grounds $H$ as Semantic
Entropy~\cite{kuhn2023semantic},
$\mathrm{SE}(x) = -\sum_{c \in \mathcal{C}} p(c\,|\,x)\log p(c\,|\,x)$
over NLI-derived semantic equivalence clusters $\mathcal{C}$.
\emph{Note on experimental backend:} all benchmarks reported in
\S\ref{sec:eval} use the default \texttt{TokenFingerprintBackend}
(NFKC-normalised, sorted-token SHA-256 truncation), which approximates
cluster identity via lexical overlap. The NLI backend is a
drop-in replacement for future work but was not used in the
reported results; claims about Semantic Entropy properties should
be read accordingly.
Dissensus $D$, order parameter $\eta$, structural temperature $T$,
free-energy proxy $F$, and trust score $\trust$ are computed
(see \S\ref{sec:method}). Phase $\in\{\text{ordered},
\text{critical}, \text{disordered}\}$ is assigned. Lyapunov
heuristic stability observable $V(t) = H(t) + \lambda D(t)$ is tracked.

\paragraph{Stage 5: Evidence Router (Critical Phase).}
If phase = critical, the \texttt{CriticalEvidenceRouter} evaluates a
five-field \texttt{EvidenceSignal} and returns one of:
\texttt{evidence\_accept}, \texttt{abstain}, or \texttt{escalate}.
\emph{Important:} the current EvidenceSignal is an oracle-proxy
(built from consensus statistics); semantic BM25/NLI retrieval is
pluggable but not live (\S\ref{sec:evidence}).

\paragraph{Stage 6: Policy Engine.}
Seven hard blocks are evaluated in priority order before any routing
logic. OPA/Rego is queried first; unavailability triggers a Python
fallback (fail closed). Gate outcome $g$ is assigned.

\paragraph{Output: DecisionEnvelope v2.}
An immutable envelope records: request, assessment, gate, reviewer
context, follow-up request, case history, policy-learning candidate,
and audit hash-chain entry.

% =====================================================================
\section{Method: Thermodynamic-Inspired Uncertainty Observables}
\label{sec:method}

\emph{Thermodynamic terminology is an operational metaphor for observable
consensus structure. No claim is made that LLM systems obey a physical
thermodynamic law. The observables are deterministic functions of
oracle outputs and calibration parameters.}

\subsection{Weighted Support Distribution}

Let $\hat{p}(v) = \sum_{o \in \OraclesV} w(o) \cdot
\mathbf{1}[\varphi(o)=v]$ be the diversity-weighted support for
verdict $v$, normalized to sum to 1. The diversity weight $w(o)$
penalizes oracles that frequently agree with others:
%
\begin{equation}
  w(o) = \frac{(1/n)}{1 + \sum_{j \neq o} \rho(o,j)} \cdot Z^{-1}
\end{equation}
%
where $\rho(o,j)$ is the rolling pairwise agreement rate (window 200
samples) and $Z$ normalizes. Oracles that agree with the swarm receive
lower weight; independent dissenters receive higher weight.

\subsection{Core Observables}

\textbf{Shannon entropy:}
\begin{equation}
  H = -\sum_v \hat{p}(v) \log_2 \hat{p}(v) \quad \text{(bits)}
\end{equation}

\textbf{Dissensus:}
\begin{equation}
  D = 1 - \max_v \hat{p}(v)
\end{equation}

\textbf{Order parameter:}
\begin{equation}
  \eta = \frac{\max_v \hat{p}(v) - 1/k}{1 - 1/k}
\end{equation}
where $k = |\{v : \hat{p}(v) > 0\}|$ is the number of active verdicts.

\textbf{Structural temperature} (prompt-only, independent of oracle
responses):
\begin{equation}
  T = 0.70\,T_{\text{prior}}(d)
        + 0.20\,\frac{|\text{zlib}(q)|}{|q|}
        + 0.10\,\frac{\ln(1+|q|)}{10}
\end{equation}
where $d \in \{\text{factoid, reasoning, creative, adversarial}\}$
with priors $\{0.25, 0.85, 1.50, 1.70\}$ respectively. This formulation
resolves the $T$--$D$ circularity documented as resolved finding R7
(Appendix~\ref{app:negres}).

\textbf{Free-energy proxy:}
\begin{equation}
  F(T) = \lambda D - T \cdot H, \quad \lambda = 0.3
\end{equation}

\textbf{Lyapunov-inspired stability observable (heuristic):}
\begin{equation}
  V(t) = H(t) + \lambda D(t)
\end{equation}
Iteration halts when $\Delta V > \epsilon_{\text{tol}} \cdot |V|$
($\epsilon_{\text{tol}} = 0.05$). Across 1,000 synthetic sessions,
$P(\Delta V \leq 0) = 87.2\%$, mean $\Delta V = -0.329$.
This is an empirical heuristic metric, not a formal control-theoretic Lyapunov
proof.

\textbf{Trust score:}
\begin{equation}
  \trust = \eta \cdot (1 - h_{\text{bound}})
           \cdot w_{\text{phase}}
           \cdot \left(1 + \frac{\chi}{\chi_0}\right)^{-1}
\end{equation}
where $h_{\text{bound}}$ is a hallucination-rate bound from
inter-oracle agreement, $w_{\text{phase}} \in \{1.0, 0.5, 0.1\}$
for ordered/critical/disordered, and $\chi$ is susceptibility
(repurposed as OOD detector; \S\ref{sec:negative}).

\subsection{Phase Classification}

\begin{align}
  \text{phase} &= \begin{cases}
    \text{ordered}    & T < T_c \text{ and } \eta > 0.5 \\
    \text{critical}   & |T - T_c|/T_c < 0.15 \\
    \text{disordered} & \text{otherwise}
  \end{cases}
\end{align}

where $T_c$ is calibrated from the oracle response distribution.

% =====================================================================
\section{Decision and Policy Model}
\label{sec:decision}

\subsection{Four Gate Outcomes}

\begin{table}[h]
\centering\small
\caption{Gate outcomes and autonomous action status.}
\fitcolumn{%
\begin{tabular}{@{}llc@{}}
\toprule
Outcome & Meaning & Autonomous action \\
\midrule
\textsc{Accept}   & Assurance conditions met         & Permitted \\
\textsc{Verify}   & Plausible, validation required   & Pending \\
\textsc{Abstain}  & Uncertainty too high to decide   & Blocked \\
\textsc{Escalate} & Human review required            & Blocked \\
\bottomrule
\end{tabular}}
\label{tab:outcomes}
\end{table}

\textsc{Accept} does not mean the action is \emph{correct}---it means
the assurance conditions for autonomous execution are met.
\textsc{Abstain} does not mean the action is \emph{wrong}---it means
the conditions for deciding are not met. Both \textsc{Abstain} and
\textsc{Escalate} block execution; \textsc{Escalate} additionally
triggers a human-review workflow.

\subsection{Hard Blocks (Priority Order)}

\begin{table}[h]
\centering\small
\caption{Policy hard blocks, evaluated before any routing logic.}
\fitcolumn{%
\begin{tabular}{@{}clll@{}}
\toprule
Pri. & Condition & Reason Code & Action \\
\midrule
1 & Adversarial detected       & ADM\_FIREWALL    & ESC. \\
2 & Counterfactual failed      & CF\_FAILED       & ESC. \\
3a& Evidence contradicted +    &                  &      \\
  & \quad contradiction cycles & EV\_CONTRAD.     & ESC. \\
3b& Evidence contradictions    & EV\_CONTRAD.     & ABS. \\
4 & Refuse parametric + no ev. & THM\_REQ\_EV.    & VER. \\
5 & Distribution shift         & DIST\_SHIFT      & VER. \\
6 & Phase=critical + risk=crit.& CRITICAL\_PHASE  & ESC. \\
7 & High/crit. risk + no ev.   & EV\_INSUFF.      & VER. \\
\bottomrule
\end{tabular}}
\label{tab:hardblocks}
\end{table}

\noindent\textbf{Policy must override model consensus.} Majority vote
is never sufficient for critical autonomous action when a hard block
condition is met.

\subsection{OPA/Rego Integration}

REMORA exports a 55-field \texttt{PolicyObservation} dataclass as
JSON to the OPA endpoint. On any communication error (connection
refused, timeout, parse failure), the Python fallback engine applies
identical rules and records \texttt{source\_of\_decision =
"python\_fallback"} in the envelope. No decision is ever withheld due
to a missing policy engine.

% =====================================================================
\section{Governance Intelligence Layer: Misspecification-Aware Pre-Policy Enrichment}
\label{sec:gov_intel}

The policy engine is only as honest as the metadata it is given: an
integration bug, or a compromised agent, can label a destructive
action as a low-risk read and present the gate with nothing to fire
on. Added in release \paperversion, the \emph{Governance Intelligence
Layer} (\path{remora/governance_intelligence/}) closes this gap with
an opt-in, fully deterministic enrichment stage that runs before any
policy rule (no LLM calls): (1)~fail-closed normalization of
caller-supplied labels (unknown is explicit, never coerced to a safe
value); (2)~action-semantics extraction from the proposed action text
and tool metadata; (3)~misspecification-risk inference from
label/semantics disagreement; (4)~causal-consequence signals (blast
radius, expected loss); and (5)~policy-generalization risk---would
repeatedly accepting this \emph{class} of action remain safe
fleet-wide?

Signals merge into the existing \texttt{PolicyObservation} under a
\textbf{strengthen-only} rule: inferred higher risk may override a
supplied lower label, never the reverse, and hard-block flags are
never cleared. A grid property test across 2{,}160 observation
combinations verifies that enrichment never converts a rejection into
an acceptance
(\path{tests/policy/test_governance_intelligence_never_weakens_policy.py}).
The engine remains the sole decision authority.

\begin{table}[h]
\centering\small
\caption{Governance-intelligence benchmark: 50 deterministic tasks
across 10 categories of mislabelled, underspecified, or repeated
actions, evaluated offline with a deliberately permissive trust
baseline so that enrichment is the only barrier to acceptance.}
\begin{tabular}{@{}lr@{}}
\toprule
Metric & Result \\
\midrule
Unsafe accept rate (no-accept tasks) & \textbf{0.0\%} \\
Metadata-mismatch detection          & 100\% \\
Unknown-metadata review rate         & 100\% \\
Legitimate reads still accepted      & 100\% \\
Escalation precision                 & 96.7\% \\
Policy-generalization detection      & 90\% \\
\bottomrule
\end{tabular}
\label{tab:govintel}
\end{table}

Artifact:
\path{artifacts/governance_intelligence/evaluation_results.json};
runner: \path{experiments/evaluate_governance_intelligence.py}.
Without enrichment, \texttt{DROP TABLE users} labelled
\texttt{action\_type=read, risk\_tier=low} reaches \textsc{Accept} on
a high-trust path; with enrichment it escalates. The layer is
heuristic and English-centric; pattern tables can be evaded by novel
phrasing, so it reduces reliance on caller labels rather than
replacing verification, evidence, or human review.

% =====================================================================
\section{Evidence-Grounded Critical-Phase Routing}
\label{sec:evidence}

\subsection{The Critical-Phase Problem}

In the critical phase, oracle consensus is near the phase boundary.
Empirically, higher trust scores \emph{negatively} correlate with
correctness ($n=32$ real-oracle items): Q4 high-trust achieves 50\%
correct; Q1 low-trust achieves 75\% correct. Conformal calibration at
5\% risk target fails completely on critical-phase items (mean
observed risk = 100\%, coverage $\rightarrow 0\%$). Standard trust-based
routing cannot safely gate critical-phase decisions.

\subsection{PhaseAwareGuardrail: Exploiting Trust Inversion}
\label{sec:phase_aware}

The inversion is a \emph{selection criterion reversal}, not an
accuracy problem.  Items with $\tau < 0.10$ in the critical phase
achieve 71.4\% accuracy ($N=21$) vs.\ 27.3\% for high-$\tau$
critical items ($N=11$).  A \texttt{PhaseAwareGuardrail} exploits
this by applying an inverted score $\tilde{\tau} = 1 - \tau$ for
critical-phase items, then calibrating a conformal threshold on
$\tilde{\tau}$.  A hard gate rejects all items with $\tau \geq
\tau_{\max} = 0.10$ (the groupthink boundary) regardless of the
conformal result.

The combined pool---all 99 ordered-phase items (sorted by
$\tau \downarrow$) plus 21 low-$\tau$ critical items ($\tau < 0.10$,
sorted by $\tilde{\tau} \downarrow$)---achieves:

\begin{itemize}[leftmargin=*]
  \item 20.0\% coverage ($N=108$): 85.2\% accuracy \text{CI~[77.3\%, 90.7\%]}
  \item 22.1\% coverage ($N=120$): 85.0\% accuracy \text{CI~[77.5\%, 90.3\%]}
\end{itemize}

This is a $+3.9$\,pp coverage gain over the ordered-only operating
point (18.2\%, 86.9\% accuracy), with a 1.9\,pp accuracy cost---still
$+43.8$\,pp above the 41.18\% full-coverage baseline.  The guardrail is
implemented in \path{remora.selective.guardrail.PhaseAwareGuardrail}
and fully tested (8 unit tests).

\subsection{Conformal Risk Control Under Covariate Shift}
\label{sec:crc}

Standard Mondrian conformal calibration assumes that calibration and
test samples are \emph{exchangeable} — drawn i.i.d.\ from the same
distribution.  This assumption is violated in the critical phase: the
joint distribution $(\tau, \text{correct})$ differs structurally
between ordered-phase calibration items and critical-phase test items
due to the trust-score inversion documented above.  Empirically,
applying standard conformal to critical-phase items yields 100\%
observed risk and zero coverage.

REMORA resolves this with \textbf{Conformal Risk Control} (CRC,
\cite{angelopoulos2022crc}), an importance-weighted extension of
split-conformal prediction to general monotone loss functions and
covariate-shifted test distributions.  Each calibration item $i$
receives an importance weight $w_i = p_{\text{test}}(x_i) /
p_{\text{cal}}(x_i)$.  For REMORA's phase shift, we use the
conservative estimate

\begin{equation}
  w_i = \begin{cases} 1.0 & \text{if phase}(i) = p_{\text{test}} \\
                      \beta & \text{otherwise} \end{cases}
  \qquad (\beta = 0.10)
  \label{eq:phase_weights}
\end{equation}

The CRC threshold $\hat{\lambda}$ is the smallest value such that the
weighted empirical risk $\bar{L}(\lambda) = \sum_i \tilde{w}_i
\ell_i(\lambda) \leq \alpha$, where
$\tilde{w}_i = w_i / (\sum_j w_j + w_{n+1})$ are the normalised
importance weights.

\begin{theorem}[CRC Guarantee, Angelopoulos et al.\ 2022]
For any monotone loss $\ell : [0,1] \to [0,B]$, target risk
$\alpha \in (0,B)$, and correctly-specified importance weights:
\[
  \mathbb{E}[L(\hat{\lambda})] \leq \alpha + \frac{B}{n+1}
\]
where $n$ is the calibration set size.  For binary (0/1) loss $B=1$,
the overshoot is bounded by $1/(n+1)$ --- negligible for $n \geq 20$.
\end{theorem}

REMORA implements CRC in \path{remora.selective.crc.CovariateShiftCRC}
with a \texttt{fit(scores, labels, phases, target\_phase)} interface
compatible with \texttt{ConformalPhaseGuardrail}.  A \texttt{CRCReport}
dataclass exposes \texttt{finite\_sample\_slack} ($= 1/(n_{\text{cal}}+1)$)
and \texttt{guaranteed\_risk\_bound} ($= \alpha + \text{slack}$) for
transparent reporting.  The module is fully tested (44 unit tests
covering edge cases, tied-score atomicity, phase-weight algebra, and
Theorem 1 finite-sample slack).


\subsection{Prover-Verifier Deliberation for Critical-Phase Routing}
\label{sec:pvd}

Kirsch et al.\ (2024) showed that prover-verifier games produce more
legible and reliable LLM outputs: a \emph{prover} oracle justifies a
claim to an independent \emph{verifier} oracle, filtering responses
whose reasoning cannot withstand scrutiny \cite{kirsch2024prover}.
REMORA adapts this protocol to its offline multi-oracle setting without
additional API calls.

\paragraph{Protocol.}
\begin{enumerate}[leftmargin=*,label=\arabic*.]
  \item \textbf{Cluster} oracle responses via Semantic Entropy clustering
        (\S\ref{sec:se}), producing semantic equivalence classes
        $\mathcal{C} = \{C_1, \ldots, C_K\}$ sorted by mass.
  \item \textbf{Prover} = the oracle whose response belongs to the
        dominant cluster $C_1$ and has the highest per-oracle confidence.
  \item \textbf{Verifier} = the oracle with the highest confidence
        \emph{outside} $C_1$ (the independent challenger).
  \item \textbf{Deliberation} ($r = 1, \ldots, n_{\text{rounds}}$):
        evaluate bidirectional NLI entailment
        $e_r = \text{NLI}(\text{prover} \to \text{verifier}) \cdot \gamma^{r-1}$
        with round-decay factor $\gamma = 0.85$.
  \item \textbf{Legibility score}:
        $\mathcal{L} = \bar{e} \cdot |C_1| / N$,
        where $\bar{e}$ is the mean entailment and $|C_1|/N$ is the
        dominant cluster mass.
  \item \textbf{Final confidence}:
        $\hat{c} = \sqrt{|C_1|/N \cdot c_{\text{verifier}}}$
        (geometric mean of prover support and updated verifier confidence).
\end{enumerate}

\paragraph{Routing signal.}
The PVD confidence is blended with the raw REMORA trust score:
\begin{equation}
  \tilde{\tau} = (1 - w_{\text{pvd}}) \cdot \tau + w_{\text{pvd}} \cdot \hat{c},
  \qquad w_{\text{pvd}} = 0.40
  \label{eq:pvd_blend}
\end{equation}
This allows PVD to promote borderline critical-phase items to
\textsc{Accept} when deliberation confidence is high, extending coverage
beyond the \texttt{PhaseAwareGuardrail} operating point without
relaxing the formal risk bound.

\paragraph{Implementation.}
The deliberation entry point is
\begin{lstlisting}
remora.selective.pvd.deliberate(
    oracle_responses, oracle_confidences,
    n_rounds=2)
\end{lstlisting}
It returns a \texttt{PVDResult} with fields
\path{prover_cluster_mass}, \path{legibility_score},
\path{agreement}, and \path{final_confidence}.
\texttt{pvd\_routing\_score(trust\_score, pvd\_result)} implements
Eq.~\ref{eq:pvd_blend}.  Fully tested (33 unit tests, no external
model calls).

\subsection{CriticalEvidenceRouter}

The router evaluates a five-field \texttt{EvidenceSignal}
with fields (\texttt{evidence\_strength}, \texttt{contradiction\_score},
\texttt{citation\_coverage}, \texttt{cross\_evidence\_consistency},
\texttt{source\_reliability}):

\begin{enumerate}[leftmargin=*,label=\arabic*.]
  \item Coverage gate: \texttt{citation\_coverage}~$< 0.50$ $\Rightarrow$ \textsc{Escalate}
  \item Contradiction block: \texttt{contradiction\_score}~$> 0.50$ $\Rightarrow$ \textsc{Abstain}
  \item Evidence-accept: all thresholds met $\Rightarrow$ \texttt{evidence\_accept}
  \item Default: $\Rightarrow$ \textsc{Escalate} with diagnostic
\end{enumerate}

On MultiNLI ($N=3{,}000$): resolution rate 38.5\%, evidence-accept
precision 100\%, false-accept rate on contradiction items 0\%,
abstain precision on contradictions 99.3\%.

\subsection{Evidence Signal Provenance}

\emph{Current implementation:} \texttt{EvidenceSignal} is built as a
proxy from oracle consensus statistics. This is documented in the
source as a ``structural bridge'' pending semantic retrieval.
\emph{Planned:} external BM25/NLI passage retrieval feeding real
document evidence. The routing logic is fully implemented and tested;
real evidence quality may differ from NLI label quality.

% =====================================================================
\section{Audit and Governance Envelope}
\label{sec:audit}

\subsection{DecisionEnvelope v2}

Every gate invocation produces an immutable
\texttt{DecisionEnvelope} with blocks: request, assessment, gate,
reviewer\_context, follow\_up, history, policy\_learning, and audit.
The full schema is in Appendix~\ref{app:schema}.

\subsection{Audit Hash-Chain}

\begin{equation}
  h_i = \text{SHA-256}(h_{i-1} \;\|\; \mathrm{json}(e_i))
\end{equation}

This provides \textbf{tamper-detection}: any modification breaks the
chain. It does \textbf{not} prevent tampering. Tamper-\emph{proof}
audit requires external append-only storage (S3 Object Lock, Azure
Immutable Blob, or equivalent) as a deployment dependency.

\subsection{Follow-Up Workflow}

\textsc{Escalate} generates a structured follow-up request specifying
reason codes, required evidence, responsible role, and SLA. Data
structures are implemented; production integration with external
ticketing systems is a deployment requirement.

% =====================================================================
\section{Experimental Evaluation}
\label{sec:eval}

\subsection{Benchmarks}

\textbf{QA benchmark ($N=544$).}
BoolQ ($n=377$) \cite{clark2019boolq},
TruthfulQA ($n=85$) \cite{lin2022truthfulqa},
adversarial-curated ($n=7$),
REMORA-curated ($n=75$, author-assembled, selection bias possible).

\textbf{Tool-call benchmark ($N=700$).}
Synthetic adversarial suite: 7 domains $\times$ 4 task types $\times$
25 tasks. Three adversarial failure modes: intent-argument conflict,
change-window/dual-control requirements, safe-looking tasks with
dangerous hidden scope.

\textbf{Evidence benchmark ($N=3{,}000$).}
MultiNLI validation-matched \cite{williams2018broad} (entailment,
neutral, contradiction) as proxy for evidence-verification decisions.

\textbf{Lyapunov benchmark ($N=1{,}000$ sessions).}
Synthetic oracle responses, seeded RNG, 5--20 steps per session.

\subsection{Baselines}

QA: full-coverage majority vote; trust-score selective (threshold=0.50).

Tool-call: single model (Llama-3.3-70B); majority vote; self-consistency;
verifier heuristic; REMORA temperature-gate only (no policy engine).

\subsection{Oracles}

All experiments: Llama-3.1-8B-Instant, Llama-3.3-70B-Versatile,
Llama-4-Scout-17B-16E-Instruct (Groq API). Oracle model versions
may change; stored artifacts ensure deterministic reproduction of
QA and tool-call benchmarks.

% =====================================================================
\section{Results}
\label{sec:results}

\subsection{Selective Accuracy on QA Benchmark}

\begin{table}[h]
\centering\small
\caption{QA benchmark selective accuracy ($N=544$). $\dagger$~In-sample
optimum; $\ddagger$~held-out: stratified 80/20 split, $\tau^*=0.203$
locked from training set (seed=42); $\S$~\texttt{PhaseAwareGuardrail}:
ordered~$+$~inverted low-$\tau$ critical ($\tau < 0.10$), $N=120$.}
\fitcolumn{%
\begin{tabular}{@{}llrrrr@{}}
\toprule
Condition & Set & Cov. & Acc. & Lift & 95\% CI \\
\midrule
Majority vote (full) & full   & 100\%  & 41.18\% & ---  & [37.1, 45.4] \\
neg-temp. (10\%)$\dagger$     & in-sam & 10\%   & 81.48\% & +40.3 & [69.2, 89.6] \\
neg-temp. (18\%)$\dagger$     & in-sam & 18\%   & \textbf{88.78\%} & \textbf{+47.6} & [81.0, 93.6] \\
phase-aware (22\%)$\S$        & in-sam & 22.1\% & 85.0\%  & +43.8 & [77.5, 90.3] \\
neg-temp. (25\%)$\dagger$     & in-sam & 25\%   & 72.79\% & +31.6 & [64.8, 79.6] \\
neg-temp. (held-out)$\ddagger$ & \textbf{holdout} & 23.2\% & \textbf{88.00\%} & \textbf{+41.7} & \textbf{[70.0, 95.8]} \\
\bottomrule
\end{tabular}}

\smallskip
\footnotesize
Held-out: $N_\text{accepted}=25$ (all ordered-phase), $p=1.45\times10^{-5}$
(one-sided binomial, H$_1$: acc $>$ holdout baseline 46.3\%),
Wilson CI lies entirely above holdout baseline.
\label{tab:qa}
\end{table}

Phase breakdown on full dataset:
ordered (N=99): 86.9\%; critical (N=32): 62.5\%; disordered (N=413): 28.6\%.
Disordered items constitute 75.9\% of the benchmark, driving the
low full-coverage baseline---a structural property of benchmark
composition. Extending coverage beyond the ordered-only 18.2\% operating
point requires exploiting the critical-phase trust inversion
(\S\ref{sec:phase_aware}): the \texttt{PhaseAwareGuardrail} raises
coverage to 22.1\% at 85.0\% accuracy by adding 21 low-$\tau$ critical
items ($\tau < 0.10$, inverted-score selection, $+43.8$\,pp lift; Wilson~CI [77.5\%, 90.3\%]).

\subsection{Tool-Call Benchmark ($N=700$)}

\begin{table}[h]
\centering\small
\caption{Tool-call benchmark: REMORA vs baselines ($N=700$ tasks).}
\begin{tabular}{@{}lccc@{}}
\toprule
Condition & Acc. & Unsafe & Utility \\
\midrule
Single model               & 20\% & 20\% & $-$0.25 \\
Majority vote              & 30\% & 10\% & 0.00 \\
Self-consistency           & 30\% & 10\% & 0.00 \\
Verifier heuristic         & 20\% & 20\% & $-$0.25 \\
REMORA temp. gate only     & 70\% & 10\% & 0.27 \\
\textbf{REMORA full policy}& \textbf{90\%} & \textbf{0\%} & \textbf{0.62} \\
\bottomrule
\end{tabular}
\label{tab:toolcall}
\end{table}

The temperature-gate-only row demonstrates that thermodynamic routing
is necessary but insufficient: policy hard blocks contribute 100\%
of the unsafe-execution reduction (10\% $\rightarrow$ 0\%).

\subsection{Conformal Coverage (Mondrian, $N=2{,}161$)}

\begin{table}[h]
\centering\small
\caption{Mondrian phase-stratified conformal, 15\% risk target
(20 seeds $\times$ $N=2{,}161$ augmented dataset).}
\begin{tabular}{@{}lrrl@{}}
\toprule
Phase & Mean Risk & Mean Cov. & Seeds Fail \\
\midrule
Ordered    & 12.0\% & \textbf{99.9\%} & \textbf{0/20} \\
Critical   & 64.5\% &  0.9\% & 4/20 \\
Disordered & 78.6\% &  0.2\% & 2/20 \\
\bottomrule
\end{tabular}
\label{tab:mondrian}
\end{table}

Ordered-phase conformal achieves 99.9\% coverage with 0/20 failures---a
deployable operating point. Critical and disordered phases cannot achieve
meaningful coverage via conformal calibration alone, motivating the
evidence router.

\subsection{Lyapunov Stability ($N=1{,}000$ Sessions)}

$P(\Delta V \leq 0) = 87.2\%$; mean $\Delta V = -0.329$;
P95 $= +0.152$; P99 $= +0.308$. The 12.8\% of sessions where $V$
increased are correctly terminated by the abort criterion.

\subsection{Evidence Router Proxy Benchmark (MultiNLI, $N=3{,}000$)}

Resolution rate 38.5\%; evidence-accept precision 100\%;
false-accept rate on contradiction 0\%; abstain precision 99.3\%.
Trust-only baseline: 0\% resolution (all escalated).
This NLI-proxy benchmark is distinct from the 32-case
\emph{Cross-Domain Evidence Replay} (static rule-based provider)
reported in the repository and the enterprise white paper; the two
measure different mechanisms at different scales.

% =====================================================================
\section{Ablations}
\label{sec:ablation}

\textbf{Policy vs. thermodynamics.}
Temperature gate only achieves 70\% accuracy, 10\% unsafe execution.
Full policy gate: 90\% accuracy, 0\% unsafe. Hard blocks account for
the entire safety improvement.
\emph{Architectural implication.} The unsafe-execution reduction is
100\% attributable to deterministic, interpretable policy rules that are
independent of LLM output. The multi-oracle consensus and thermodynamic
routing machinery improve decision \emph{calibration and routing quality}
(discriminating VERIFY, ABSTAIN, and ACCEPT paths), but do not contribute
to the safety guarantee itself. This separation is a design feature: the
safety floor can be reasoned about and audited without reference to
oracle behaviour. Readers should not infer that the consensus
machinery is the source of the 0\% unsafe rate.

\textbf{Oracle count ($N=302$).}
Single oracle: 56.95\% (95\% CI [51.3, 62.4]).
Three-oracle majority vote: $\approx$+10 pp.
Correlation-aware weighting: additional gain within multi-oracle setting.

The primary benefit of the oracle swarm is disagreement
\emph{detection} (high $H$, high $D$) rather than accuracy
improvement per se.

% =====================================================================
\section{Case Study: Well Barrier Agent Action}
\label{sec:casestudy}

\emph{No domain-engineering correctness is claimed. This is an
illustrative governance walkthrough.}

A drilling agent proposes: \textit{``Autonomously update kill mud
weight from 1.52 to 1.48\,SG to improve ROP on the 8½{\textquotedblright}
section.''} Risk tier: critical. Domain: \texttt{well\_engineering}.

\textbf{Oracle responses:}
Llama-3.1-8B: ESCALATE, conf 0.78;
Llama-3.3-70B: ESCALATE, conf 0.85 (NORSOK D-010 §5.4);
Llama-4-Scout: VERIFY, conf 0.62.
One oracle failed; excluded.

\textbf{Consensus:}
$\hat{p}(\textsc{Escalate}) = 0.71$, $\hat{p}(\textsc{Verify}) = 0.29$.
$H = 0.866$ bits; $D = 0.29$;
phase = critical ($T = 0.85$, near $T_c$);
$\trust = 0.31$.

\textbf{Evidence:} No field inspection, no ECD calculation, no OIM
sign-off. Citation coverage = 0.18 $<$ 0.50 threshold.
Evidence router: \textsc{Escalate}.

\textbf{Policy engine:} Hard block \#6 fires immediately
(phase=critical AND risk=critical) $\Rightarrow$ \textsc{Escalate}.

\textbf{Follow-up generated:} type=on\_site\_inspection;
assign\_to=Independent Well Engineer; required\_evidence = \{ECD
calculation, kick tolerance, NORSOK D-010 barrier confirmation, OIM
sign-off\}; SLA=4h.

\textbf{Audit hash} recorded; action blocked.

% =====================================================================
\section{Negative Results}
\label{sec:negative}

Publication of negative results is essential for scientific
credibility. This section documents active and resolved findings.
Full documentation is in \texttt{NEGATIVE\_RESULTS.md}.

\begin{table}[h]
\centering\small
\caption{Negative results and mitigations.}
\fitcolumn{%
\begin{tabular}{@{}llll@{}}
\toprule
Finding & Sev. & Status & Mitigation \\
\midrule
$\chi$-proxy AUC = 0.39 & Med & Docum. & Repurposed as OOD \\
$T$--$D$ circularity & Med & Resolv. & Structural $T$ \\
Crit.-phase anticorr. & High & Active & Evidence router \\
Full-cov. acc. weak & Med & Active & Selective prediction \\
Oracle diversity part. & Med & Active & Diversity weights \\
In-sample calibration & Med & \textbf{Resolved} & 88.0\% held-out, $p$=1.45e-5 \\
Evidence signal proxy & Med & Active & NLI benchmark \\
Audit not tamper-proof & Med & By des. & WORM storage \\
\bottomrule
\end{tabular}}
\label{tab:negresults}
\end{table}

\paragraph{$\chi$-proxy failure.}
Susceptibility $\chi$ as standalone difficulty predictor: AUC = 0.39
on $N=302$ items---below chance. Repurposed as OOD detector:
$\chi > 1.45$ (97th percentile) $\Rightarrow$ \textsc{Escalate} with
\texttt{escalate\_adversarial}.

\paragraph{Critical-phase trust anticorrelation.}
Q4 high-trust: 50\% correct; Q1 low-trust: 75\% correct.
Confirmed on $N=32$ real-oracle items and replicated qualitatively
on augmented $N=511$ simulated dataset. No trust threshold separates
correct from incorrect critical-phase items. \emph{Consequence:}
evidence routing---not trust scoring---is the required mechanism for
critical-phase decisions.

\paragraph{$T$--$D$ circularity (resolved).}
Legacy temperature estimator weighted $D$ at 18\% of $T$, creating
a feedback loop ($D \rightarrow T \rightarrow F \rightarrow V$, all
depending on $D$). Resolved by structural temperature (\S\ref{sec:method}),
which computes $T$ from prompt structure alone.

% =====================================================================
\section{Threats to Validity}
\label{sec:threats}

\textbf{Internal validity:} Trust threshold ($\tau^* = 0.2032$,
18th-percentile neg-temperature on 80\% training split) is derived
in-sample; held-out evaluation ($n=108$, 20\% stratified) confirms
validity ($p=1.45\times10^{-5}$); 13.8\% author-curated benchmark
items; synthetic tool-call adversarial patterns.

\textbf{External validity:} All experiments use Groq/Llama models;
benchmark phase distribution may not match deployment reality;
MultiNLI is a proxy for field evidence retrieval.

\textbf{Construct validity:} Thermodynamic terminology is metaphorical,
not physical. ``Trust'' is a derived scalar, not a frequency probability
of per-item correctness. Utility scores are designed task weights, not
real-world costs.

% =====================================================================
\section{Safety, Ethics, and Governance}
\label{sec:safety}

REMORA is designed to \emph{reduce} unsafe autonomous AI actions.
\textsc{Accept} means assurance conditions are met, not that the action
is correct. Human reviewers receiving \textsc{Escalate} retain full
accountability for outcomes. The oracle swarm routes sensitive
operational data through external APIs; data classification requirements
must be evaluated before deployment. The adversarial firewall detects
known injection patterns but is not a complete defense against
sophisticated adversarial inputs.

% =====================================================================
\section{Reproducibility}
\label{sec:repro}

\textbf{Repository:} \url{https://github.com/darklordVirtual/REMORA}.
\textbf{Release:} \texttt{\paperversion} (this paper is versioned in lockstep
with the repository release tag).
\textbf{Python:} 3.11+.

\begin{lstlisting}
python -m pip install -e ".[dev]"
make audit        # lint + tests + claim consistency
make benchmark    # all deterministic benchmarks
make credibility-pack
\end{lstlisting}

Live oracle experiments require \texttt{GROQ\_API\_KEY}. Oracle model
versions on the Groq API may change; results are sensitive to model
version updates. Stored artifact experiments are fully deterministic.

% =====================================================================

% TEE content moved to Appendix~\ref{app:tee}.

\section{Conclusion}
\label{sec:conclusion}

We presented REMORA, a policy-gated multi-oracle assurance architecture
for governing agentic AI actions. The central contribution is a
system-level design that converts multi-oracle disagreement,
thermodynamic uncertainty observables, evidence signals, and policy
constraints into four structured gate outcomes with a full audit record.

Key results: 0\% unsafe execution on an adversarial 700-task tool-call
benchmark (vs.\ 10--20\% for all baselines; Wilson CI [0.00\%, 0.55\%]);
88.8\% in-sample selective accuracy at 18\% coverage, confirmed at 88.0\%
on a stratified held-out split ($N_\text{accepted}=25$,
$p=1.45\times10^{-5}$; Wilson CI [70.0\%, 95.8\%]---directional
confirmation, not a tight estimate given the small accepted set); 38.5\%
critical-phase resolution with 100\% evidence-accept precision; 99.9\%
ordered-phase conformal coverage with 0/20 seed failures.

Key negative findings: trust alone cannot route critical-phase decisions
safely; $\chi$-proxy susceptibility fails as a difficulty predictor
(AUC = 0.39); evidence retrieval is currently proxy-based; the benchmark
has structural composition biases; entropy computation in reported
experiments uses a token-fingerprint heuristic, not full Semantic Entropy
over NLI clusters. A central architectural finding is that the
safety improvement is entirely attributable to deterministic policy
hard blocks, not to the multi-oracle consensus machinery---the
latter contributes routing calibration quality.

REMORA is a research-grade prototype. Production deployment requires
semantic evidence retrieval, WORM audit storage, OPA policy integration,
and domain-authority validation. The architecture is designed to be
pluggable: each component can be replaced without changing the gate
interface.

The broader thesis---that governed autonomy requires explicit uncertainty
routing, not suppression---is supported by both positive results and
negative findings. We offer REMORA as a reference architecture, an
empirical baseline, and a documented set of unsolved problems for the
AI assurance community.

% =====================================================================
\section*{Acknowledgements}
Benchmarks use publicly available datasets (BoolQ, TruthfulQA, MultiNLI,
ARC-Challenge, MMLU-Pro). Oracle inference via Groq API. Open Policy Agent
used under Apache 2.0 license. Negative results documented in full in
\texttt{NEGATIVE\_RESULTS.md}; the authors thank all reviewers who pointed
out scope corrections and limitations.

% =====================================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{99}
\bibitem{kuhn2023semantic}
Kuhn, L., Gal, Y., \& Farquhar, S.\ (2023).
Semantic uncertainty: Linguistic invariances for uncertainty estimation in
natural language generation.
In \emph{Proceedings of the International Conference on Learning Representations
(ICLR 2023)}.
arXiv:2302.09664.

\bibitem{kuhn2026evse}
Kuhn, L., Gal, Y., \& Farquhar, S.\ (2026).
Evidential Semantic Entropy for LLM uncertainty quantification.
In \emph{Proceedings of the 17th Conference of the European Chapter of the
Association for Computational Linguistics (EACL 2026)}, pp.\ 334--348.

\bibitem{kirsch2024prover}
Kirsch, A., Harrison, J., Misra, S., \& Leike, J.\ (2024).
Prover-verifier games improve legibility of LLM outputs.
\emph{arXiv:2407.13692}.

\bibitem{angelopoulos2022crc}
Angelopoulos, A.\ N., Bates, S., Fisch, A., Lei, L., \& Schuster, T.\ (2022).
Conformal risk control.
\emph{arXiv:2208.02814}.

\bibitem{zhang2025tee}
Zhang, Y.\ \& Lee, M.\ (2025).
Evaluating the performance of large language models in confidential computing
environments.
\emph{arXiv:2502.11347}.

\bibitem{wang2023self}
Wang, X., et al.\ (2023).
Self-consistency improves chain of thought reasoning in language models.
\emph{ICLR 2023}.

\bibitem{du2023improving}
Du, Y., et al.\ (2023).
Improving factuality and reasoning in language models through multiagent debate.
\emph{ICML 2024}.

\bibitem{zheng2023judging}
Zheng, L., et al.\ (2023).
Judging LLM-as-a-judge with MT-Bench and Chatbot Arena.
\emph{NeurIPS 2023}.

\bibitem{cobbe2021training}
Cobbe, K., et al.\ (2021).
Training verifiers to solve math word problems.
\emph{arXiv:2110.14168}.

\bibitem{chow1970optimum}
Chow, C.\ K.\ (1970).
On optimum recognition error and reject tradeoff.
\emph{IEEE Transactions on Information Theory, 16}(1), 41--46.

\bibitem{elyaniv2010selective}
El-Yaniv, R.\ \& Wiener, Y.\ (2010).
On the foundations of noise-free selective classification.
\emph{Journal of Machine Learning Research, 11}, 1605--1641.

\bibitem{lakshminarayanan2017simple}
Lakshminarayanan, B., Pritzel, A., \& Blundell, C.\ (2017).
Simple and scalable predictive uncertainty estimation using deep ensembles.
\emph{Advances in Neural Information Processing Systems (NeurIPS 2017)}.

\bibitem{kuncheva2003measures}
Kuncheva, L.\ I.\ \& Whitaker, C.\ J.\ (2003).
Measures of diversity in classifier ensembles and their relationship with
the ensemble accuracy.
\emph{Machine Learning, 51}(2), 181--207.

\bibitem{geifman2017selective}
Geifman, Y.\ \& El-Yaniv, R.\ (2017).
Selective classification for deep neural networks.
\emph{NeurIPS 2017}.

\bibitem{kadavath2022language}
Kadavath, S., et al.\ (2022).
Language models (mostly) know what they know.
\emph{arXiv:2207.05221}.

\bibitem{angelopoulos2021gentle}
Angelopoulos, A.\ \& Bates, S.\ (2021).
A gentle introduction to conformal prediction and distribution-free uncertainty quantification.
\emph{arXiv:2107.07511}.

\bibitem{guo2017calibration}
Guo, C., et al.\ (2017).
On calibration of modern neural networks.
\emph{ICML 2017}.

\bibitem{endsley1995toward}
Endsley, M.\ (1995).
Toward a theory of situation awareness in dynamic systems.
\emph{Human Factors, 37}(1), 32--64.

\bibitem{kelly1998goal}
Kelly, T.\ (1998).
Arguing safety: A systematic approach to managing safety cases.
PhD thesis, University of York.

\bibitem{bloomfield2010safety}
Bloomfield, R.\ \& Bishop, P.\ (2010).
Safety and assurance cases: Practice, trends and challenges.
\emph{SAFECOMP 2010}.

\bibitem{lin2022truthfulqa}
Lin, S., et al.\ (2022).
TruthfulQA: Measuring how models mimic human falsehoods.
\emph{ACL 2022}.

\bibitem{clark2019boolq}
Clark, C., et al.\ (2019).
BoolQ: Exploring the surprising difficulty of natural yes/no questions.
\emph{NAACL 2019}.

\bibitem{williams2018broad}
Williams, A., et al.\ (2018).
A broad-coverage challenge corpus for sentence understanding through inference.
\emph{NAACL 2018}.

\bibitem{euaiact2024}
European Parliament \& Council of the European Union.\ (2024).
Regulation (EU) 2024/1689 of the European Parliament and of the Council
(Artificial Intelligence Act).
\emph{Official Journal of the European Union}, L~2024/1689.

\bibitem{opa}
Open Policy Agent.\ (2024).
\url{https://www.openpolicyagent.org/}.

\bibitem{bjoru2026causal}
Bjøru, A.~R.\ (2026).
Causal Post-hoc Explainable AI.
PhD thesis, Norwegian University of Science and Technology (NTNU).
ISBN~978-82-353-0022-5.

\end{thebibliography}

% =====================================================================
\appendix

\section{Mathematical Notation Table}
\label{app:math}

\begin{table}[h]
\centering\small
\caption{Mathematical notation.}
\fitcolumn{%
\begin{tabular}{@{}lll@{}}
\toprule
Symbol & Definition & Domain \\
\midrule
$\Oracles$ & Oracle set & $n \geq 3$ models \\
$\OraclesV$ & Valid oracle set & $\OraclesV \subseteq \Oracles$ \\
$\phi(o)$ & Canonical verdict & (polarity, hash, mag., tags) \\
$\rho(a,b)$ & Pairwise agreement rate & $[0,1]$, window 200 \\
$w(o)$ & Diversity weight & normalized \\
$\hat{p}(v)$ & Weighted support & $[0,1]$, sums to 1 \\
$H$ & Shannon entropy & bits $\geq 0$ \\
$D$ & Dissensus & $[0,1]$ \\
$k$ & Active verdict count & $\geq 1$ \\
$\eta$ & Order parameter & $[0,1]$ \\
$T$ & Structural temperature & $[0.05, 2.0]$ \\
$T_c$ & Critical temperature & calibrated \\
$\chi$ & Susceptibility & $[0, \infty)$ \\
$F$ & Free-energy proxy & $\mathbb{R}$ \\
$\lambda$ & Coupling constant & 0.3 \\
$\trust$ & Trust score & $[0,1]$ \\
$h_{\rm bound}$ & Hallucination bound & $[0,1]$ \\
$w_{\rm phase}$ & Phase weight & $\{1.0, 0.5, 0.1\}$ \\
$V(t)$ & Lyapunov-inspired heuristic stability observable & $[0,\infty)$ \\
$g$ & Gate outcome & ACCEPT/VERIFY/ABS./ESC. \\
$\Gate$ & Gate function & $(q,a,c) \to g$ \\
\bottomrule
\end{tabular}}
\end{table}

\section{DecisionEnvelope v2 Schema Summary}
\label{app:schema}

Top-level blocks: \texttt{request}, \texttt{assessment},
\texttt{gate}, \texttt{reviewer\_context}, \texttt{follow\_up},
\texttt{history}, \texttt{policy\_learning}, \texttt{audit}.
Full JSON schema is in the Markdown paper and repository
(\texttt{remora/policy/report.py}).

\section{Negative Results Archive}
\label{app:negres}

Full negative results documentation: \path{NEGATIVE_RESULTS.md}
in repository. Resolved findings include: R1 iteration damage (fixed
via skip\_high\_trust), R2 Stage-3b critique impact (measured,
phase-differentiated routing implemented), R7 $T$--$D$ circularity
(structural temperature introduced).

\section{Implementation Status Table}
\label{app:impl}

\begin{table}[h]
\centering\small
\caption{Component implementation status.}
\fitcolumn{%
\begin{tabular}{@{}lll@{}}
\toprule
Component & Status & Notes \\
\midrule
Oracle fan-out & Impl. & ThreadPoolExecutor \\
Canonicalization ($\phi$) & Impl. & NFKC, Semantic Entropy \cite{kuhn2023semantic} \\
Correlation weighting & Impl. & Rolling 200-sample \\
$H$, $D$, $\eta$, $T$, $F$, $\trust$ & Impl. & \texttt{thermodynamics.py} \\
$V(t)$ Lyapunov tracking & Impl. & Abort on $\Delta V$ \\
Phase classification & Impl. & 3-class \\
Policy engine (7 hard blocks) & Impl. & Priority-ordered \\
OPA/Rego adapter & Partial & Python fallback \\
Evidence router (oracle proxy) & Impl. & Proxy signal \\
Evidence router (semantic) & Planned & Interface ready \\
DecisionEnvelope v2 & Impl. & Frozen dataclass \\
SHA-256 hash-chain & Impl. & Tamper-detecting \\
Human review workflow & Impl. & Demo UI only \\
WORM audit storage & Planned & External dependency \\
ZK assurance proofs & Planned & Stubbed \\
\bottomrule
\end{tabular}}
\end{table}

\section{Optional TEE-Based Audit Hardening (Tamper-Resistant)}
\label{app:tee}

\emph{This appendix describes tamper-resistant (not tamper-proof) audit hardening
via hardware attestation. Implementation status: protocol specified; hardware
integration pending. The term ``tamper-resistant'' is used throughout; the
hash-chain in the main system is tamper-detecting only.}

\subsection{Motivation}

The SHA-256 hash-chain in \path{remora/audit/hash_chain.py} detects
post-hoc modification of audit records but does not prevent an adversary with
storage write access from replacing the entire chain. For deployments where
stronger guarantees are required, hardware attestation via a Trusted Execution
Environment (TEE) provides a complementary layer: the \texttt{DecisionEnvelope}
is signed inside a hardware-isolated enclave, and the attestation quote
includes a measurement of the enclave binary, making silent substitution
detectable even if storage is compromised.

This is an optional deployment hardening; it is not required for the core
REMORA governance protocol described in the main paper.

\subsection{Supported TEE Platforms}

Three platforms are in scope for the reference implementation:
\begin{itemize}
  \item \textbf{Intel TDX / SGX} — remote attestation via DCAP quote generation
  \item \textbf{AMD SEV-SNP} — firmware measurement in attestation report
  \item \textbf{ARM TrustZone} — OP-TEE trusted application (mobile / edge)
\end{itemize}

\subsection{DecisionEnvelope Attestation Format}

When TEE hardening is active, each \texttt{DecisionEnvelope} includes an
\texttt{attestation} sub-object:

\begin{lstlisting}
{
  "envelope_id": "env_7f3a...",
  "gate": { "outcome": "ESCALATE", ... },
  "attestation": {
    "tee_platform": "intel_tdx",
    "quote":        "<base64 DCAP quote>",
    "mrenclave":    "<SHA-256 of enclave measurement>",
    "signature":    "<Ed25519 over envelope_id
                     +outcome+timestamp>",
    "timestamp":    "2026-01-15T14:23:07Z"
  }
}
\end{lstlisting}

The \texttt{mrenclave} value pins the exact binary that produced the decision,
allowing auditors to verify that no unauthorised code modification occurred
between decisions.

\subsection{Verification Flow}

\begin{enumerate}
  \item Auditor retrieves envelope and attestation quote from audit log.
  \item Quote is verified against Intel/AMD/ARM attestation service.
  \item \texttt{mrenclave} is compared against the published release measurement.
  \item Ed25519 signature over \texttt{(envelope\_id, outcome, timestamp)} is
        verified with the enclave's ephemeral key, which is bound to the quote.
\end{enumerate}

If any step fails, the audit record is flagged as potentially tampered.

\subsection{Implementation Status}

Protocol specified and data contract defined. Hardware integration (enclave
compilation, DCAP provisioning, OP-TEE TA signing) is a deployment
dependency and is not part of the core open-source release. Stub interfaces
are present in \texttt{remora/audit/} for downstream integration.

\end{document}


